Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-10 draft/paper.tex

\begin{definition}[The $\Set$ model of $\ToS$] For each $i$ and $j$, we have $\bA : \ToS_{\max(i+1, j)+1, j}$, which can be given by restricting the $\bM_{i,j}$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitution, types, term, and we only have
actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bA} && = \Set_{\max(i+1, j)} \\
  & \Sub_{\bA}\,\Gamma\,\Delta && = \Gamma \ra \Delta\\
  & \Ty_{\bA}\,\Gamma && = \Gamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bA}\,\Gamma\,A && = (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for some $\Gamma : \Con_{\syn_j}$, the type of $\Gamma$-algebras at level
$i$ is given by $\llb\Gamma\rrb_{\bA}$, where we implicitly lift $\syn_j :
\ToS_{j+1, j}$ to $\ToS_{\max(i+1,j)+1, j}$. E.g.\, $\llb\NatSig\rrb_{\bA}$
yields a left-nested $\Sigma$-type of pointed sets with an endofunction. Also,
$\llb\Gamma\rrb_{\bM}$ extends $\llb\Gamma\rrb_{\bA}$ to an flCwF of
$\Gamma$-algebras, and $\llb\Gamma\rrb_{\bA} = \Con_{\llb\Gamma\rrb_{\bM}}$.
\end{definition}

@@

We would like to show that term algebras are initial, but we want to do this
\emph{on all universe levels}, i.e.\ that terms algebras are initial when lifted
to any higher level. This requires showing that QII algebras are
cumulative. We do this by induction on syntactic signatures.

@@

Cartmell \cite{gat} defines generalized algebraic theories (GATs) using type
theoretic syntax. Compared to our QII signatures, his signatures support sort
equations but do not cover infinitary constructors or recursive equations. A way
to encode sort equations in our signatures is using isomorphisms instead of
equalities. In contrast to our algebraic definition, Cartmell's signatures are
given by presyntax, named variables and typing relations, there is no
explicit model theory provided for signatures, and no explicit term model
construction. Cartmell focuses instead on showing that contextual categories
serve as classifying categories for GATs.

@@

Streicher's initiality proof \cite{streicher2012semantics} is a direct
construction of a concrete QIIT: that of the syntax of a type theory. He uses
preterms, typing relations and quotients to construct the initial model of type
theory in a set-theoretic metalanguage. A recent formalization of the same
result for a larger type theory was done by Brunerie et al.\ \cite{brunerie} in
Agda with UIP, function extensionality and quotients. Lafont et
al.\ \cite{induction-is-enough} construct all finitary inductive-inductive types
in Agda using only UIP and function extensionality. Examples of
inductive-inductive types are constructed without using UIP by Hugunin
\cite{jasper}.

@@

Existence of the theory of signatures:

\cite[Section 9]{lumsdaineShulman} shows that infinitary QIITs are not
constructible in type theory without choice, so the existence of the
theory of signatures is a non-conservative extension of extensional
type theory. The setoid model supports an example QIT which branches
over an arbitrary setoid, and this suggests that it might support the
theory of signatures as well.

>>>>

\begin{definition}[The $\Set$ model of $\ToS$] For each $i$ and $j$, we have $\bA : \ToS_{\max(i+1, j)+1, j}$, which can be given by restricting the $\bM_{i,j}$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitutions, types, terms, and we only
have actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bA} && = \Set_{\max(i+1, j)} \\
  & \Sub_{\bA}\,\Gamma\,\Delta && = \Gamma \ra \Delta\\
  & \Ty_{\bA}\,\Gamma && = \Gamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bA}\,\Gamma\,A && = (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for some $\Gamma : \Con_{\syn_j}$, the type of $\Gamma$-algebras at level
$i$ is given by $\llb\Gamma\rrb_{\bA}$, where we implicitly lift $\syn_j :
\ToS_{j+1, j}$ to $\ToS_{\max(i+1,j)+1, j}$. E.g.\, $\llb\NatSig\rrb_{\bA}$
yields a left-nested $\Sigma$-type of pointed sets with an endofunction. Also,
$\llb\Gamma\rrb_{\bM}$ extends $\llb\Gamma\rrb_{\bA}$ to an flCwF of
$\Gamma$-algebras, and $\llb\Gamma\rrb_{\bA} = \Con_{\llb\Gamma\rrb_{\bM}}$.
\end{definition}

@@

We would like to show that term algebras are initial, but we want to do this
\emph{on all universe levels}, i.e.\ that term algebras are initial when lifted
to any higher level. This requires showing that QII algebras are
cumulative. We do this by induction on syntactic signatures.

@@

Cartmell \cite{gat} defines generalized algebraic theories (GATs) using type
theoretic syntax. Compared to our QII signatures, his signatures support sort
equations but do not cover infinitary constructors or recursive equations. A way
to encode sort equations in our signatures is using isomorphisms instead of
equalities. In contrast to our algebraic definition, Cartmell's signatures are
given by presyntax, named variables and typing relations, there is no explicit
model theory provided for signatures, and no explicit term model construction is
given. Cartmell focuses instead on showing that contextual categories serve as
classifying categories for GATs.

@@

%% Streicher's initiality proof \cite{streicher2012semantics} is a direct
%% construction of a concrete QIIT: that of the syntax of a type theory. He uses
%% preterms, typing relations and quotients to construct the initial model of type
%% theory in a set-theoretic metalanguage. A recent formalization of the same
%% result for a larger type theory was done by Brunerie et al.\ \cite{brunerie} in
%% Agda with UIP, function extensionality and quotients. Lafont et
%% al.\ \cite{induction-is-enough} construct all finitary inductive-inductive types
%% in Agda using only UIP and function extensionality. Examples of
%% inductive-inductive types are constructed without using UIP by Hugunin
%% \cite{jasper}.

@@

An important motivation of the current work was to use QIITs as a framework for
algebraic theories, with the metatheory of type theories in mind as a key
application. We would prefer QIITs to
\begin{itemize}
\item Be formally precise.
\item Not gloss over issues of size.
\item Be rich enough to cover most type theories in the wild, including the theory
      of QIIT signatures.
\item Be direct enough, so that signatures for type theories can
      be written out without excessive encoding overhead.
\item Be suitable for practical implementation in proof assistants.
\item Be reducible to a minimal set of basic type formers.
\end{itemize}

With the current work, we have improved the state of QIITs with respect to
the above criteria. However, a number of open research problems remain.

With regards to the expressiveness of QIIT signatures, we do not yet
support sort equations, i.e. equations of elements of $\Tm\,\Gamma\,U$ in
signatures. Sort equations are included in Cartmell's generalized algebraic
theories \cite{gat}, and they appear to be highly useful for giving an algebraic
representation for Russell-style universes and cumulative universes
\cite{sterling2019algebraic}. We leave this for future work, but we note that
the current isofibration-based semantics does not work in the presence of strict
sort equations, since they are not invariant under isomorphism; instead, sort
equations are compatible with the stricter semantics of
\cite{kaposi2019constructing}.

While we have made an effort to shape the syntax and semantics of QIITs to
be amenable to implementation in proof assistants, much needs to be done before
we can have a practical implementation. For one, we would need to consider QIITs
in a type theory where transports along equality proofs compute, and would need
to work out computing transports for QIITs. Cubical Agda has recently made
strides in implementing HITs \cite{vezzosicubical}, but as of now it does not
support computing transports on indexed inductive types.

With regards to the reduction of QIITs to simple type formers, the
reduction of infinitary QIITs appears to be more challenging than the finitary
case. \cite[Section 9]{lumsdaineShulman} shows that infinitary QIITs are not
constructible in type theory without choice, so such QIITs are a
non-conservative extension of extensional type theory. In the finitary case, a
generalization of the approach in \cite{induction-is-enough} seems promising ;
this amounts to a Streicher-style initial algebra construction
\cite{streicher2012semantics} for the theory of finitary QIIT signatures. In
particular, Brunerie et al. \cite{brunerie} has formalized in Agda this
construction for a comparable type theory, using UIP, function extensionality,
propositional extensionality and simple quotient types.

Another line of possible future work would be to explore a more general
functorial style of semantics for QIITs. So far, we considered set-based
1-categorical semantics, which is what we need when we want to reason
inductively about syntaxes of type theories. However, it would be fruitful to
consider algebras in structured categories other than the category of sets.
